Notes on filtrations for logics that contain {\bf K5}

### 2021-07-08 filtration-stuff/NewText.tex
bisimulation stuff

@@

@@

@@

The ${\bf K5}$-closure (an Euclidean Horn closure of a binary relation) has the following equivalent definitions:
\begin{lemma} \label{equivHorn}
  Let $\mathcal{F} = \langle W, R \rangle$ be a Kripke frame and $R \subseteq R^{\bf K5}$.
  The following conditions are equivalent:

@@

Note that the following property holds not for every ${\bf K5}^{i,j}$, consider ${\bf KB}$ as an example (clearly, ${\bf KB} = {\bf K5}^{0,1}$).

\begin{lemma}
${\bf K5}^{i,j}$ has $n$ non-equivalent sequences of modalities, if . TODO: define $n$.
\end{lemma}

@@

  \begin{enumerate}
  \item  $\mathcal{M} = \langle W, R, \vartheta \rangle$ be a ${\bf K}5$-model. Consider $\Gamma' = \Gamma \cup \{ \Diamond \Box \psi \: | \: \Box \psi \in \Gamma \}$. Let $\Delta$ be any finite and $\operatorname{Sub}$-closed extension of $\Gamma'$ that contains all $\Diamond \Box \psi$ for each $\Box \psi \in \Delta$. Then a model $\widehat{M} = \langle W / \sim_{\Delta}, (R^{min}_{\Delta})^{E}, \widehat{\vartheta} \rangle$ is a filtration of $\mathcal{M}$ through $\Delta$.
  \item Let $\mathcal{L} = {\bf K} \oplus \Diamond \Diamond \Diamond p \to \Diamond p$, then we have the similar statement for $\mathcal{L}$, where $\Delta$ is an extension of $\Gamma' = \{ \Diamond \Diamond \psi \: | \: \Diamond \psi \}$ that contains all $\Diamond \Diamond \varphi$ for $\Diamond \varphi \in \Delta$.
  \end{enumerate}
\end{lemma}

>>>>

\begin{defin}
  A normal modal logic is a set of formulas that contains all Boolean tautologies,
  formulas $\Diamond p \lor \Diamond q \leftrightarrow \Diamond (p \lor q)$ and
  $\Diamond \bot \leftrightarrow \bot$, and is closed under Modus Ponens, Substitution, and
  Monotonicity:
  from $\varphi \rightarrow \psi$ infer $\Diamond \varphi \rightarrow \Diamond \psi$.
\end{defin}

\begin{defin} An Kripke model is a triple $\mathcal{M} = \langle W, R, \vartheta \rangle$,
  where $R \subseteq W \times W$, $\vartheta : \operatorname{PV} \to 2^W$, and
  the connectives have the following semantics:

  \begin{enumerate}
    \item $\mathcal{M}, w \models p \Leftrightarrow w \in \vartheta(p)$
    \item $\mathcal{M}, w \models \neg \varphi \Leftrightarrow \mathcal{M}, w \nvDash \varphi$
    \item $\mathcal{M}, w \models \varphi \lor \psi \Leftrightarrow \mathcal{M}, w \models \varphi \text{ or } \mathcal{M}, w \models \psi$
    \item $\mathcal{M}, w \models \Diamond \varphi \Leftrightarrow \exists v \in R(w) \: \mathcal{M}, v \models \varphi$
  \end{enumerate}
\end{defin}

A modal logic $\mathcal{L}$ is called locally finite, if every fragment of $\mathcal{L}$ with a finite number of variables is finite. Equivalently, its weak canonical model generated by $k < \omega$ variables is finite. Moreover, if $\mathcal{L}$ is locally finite, then $\mathcal{L} = Log(\mathcal{F}_{\mathcal{L}} \upharpoonright k \: | \: k < \omega)$, where $\mathcal{F}_{\mathcal{L}} \upharpoonright k$ is a weak canonical frame generated by $k$ variables.

\begin{defin}
The modal depth of a formula is defined recursively by the following function:
\begin{center}
  $md(p_i) = 0$

  $md(\neg \varphi) = md(\varphi)$

  $md(\varphi \lor \psi) = \max(md(\varphi), md(\psi))$

  $md(\Diamond \varphi) = 1 + md(\psi)$
\end{center}

Let $\mathcal{L}$ be a modal logic, then the modal depth of $\mathcal{L}$, denoted as $md(\mathcal{L})$, is the minimal $n$ such that any $\varphi \in \mathcal{L}$ is equivalent to some $\psi$ such that $md(\psi) = n$. If such an $n$ does not exists, then $md(\mathcal{L}) = \infty$.
\end{defin}

\subsection{Filtrations}

@@

\subsection{Horn closure}

@@

\subsection{Bisimulations}

A model $\mathcal{M}$ is called \emph{weak} model, if its valuation is restriced to some finite subset of variables, namely $PV \upharpoonright k$, if the cardinality of this subset equals $k$.

\begin{defin}
Let $\mathcal{M}_1 = \langle W_1, R_1, \vartheta_1 \rangle$ and $\mathcal{M}_2 = \langle W_2, R_2, \vartheta_2 \rangle$ be weak Kripke models (with the same variables), $x \in W_1$, and $y \in W_2$, then $x$ and $y$ are $0$-equivalent, if the following holds:

\begin{center}
  $x \equiv_0 y$ iff for each $p \in PV \upharpoonright k$ (or for each $p \in PV$) one has $\mathcal{M}_1, x \models p \Leftrightarrow \mathcal{M}_2, y \models p$
\end{center}
\end{defin}

\begin{defin}
  Let $\mathcal{M}_1 = \langle W_1, R_1, \vartheta_1 \rangle$ $\mathcal{M}_2 = \langle W_2, R_2, \vartheta_2 \rangle$ be Kripke models, then a bisimulation between $\mathcal{M}_1$ and $\mathcal{M}_2$ is a binary relation $E \subseteq W_1 \times W_2$ between two models such that:
  \begin{itemize}
    \item (0) $x_1 E x_2$ implies $x_1 \equiv_0 x_1$
    \item ({\bf zig}) $x_1 E y_1$ and $x_1 R_1 z$ implies $y_1 R_2 y_2$ and $z E y_2$ for some $y_2 \in W_2$
    \item ({\bf zag}) $x_1 E y_1$ and $y_1 R_2 y_2$ implies $x_1 R_1 x_2$ and $x_2 E y_2$ for some $x_2 \in W_1$
  \end{itemize}
\end{defin}

Models with designated points $\langle \mathcal{M}_1, x \rangle$ and  $\langle \mathcal{M}_2, y \rangle$ are bisimilar, if there exists a bisimulation $E$ between them with $x E y$. We denote that as $\langle \mathcal{M}_1, x \rangle \: \underline{\leftrightarrow} \: \langle \mathcal{M}_2, y \rangle$. Moreover, we have
\begin{center}
  $\langle \mathcal{M}_1, x \rangle \: \underline{\leftrightarrow} \: \langle \mathcal{M}_2, y \rangle$ implies $Th(\mathcal{M}_1, x) = Th(\mathcal{M}_2, y)$.
\end{center}

\begin{defin}
  Let $\langle \mathcal{M}_1, x \rangle$ and $\langle \mathcal{M}_2, y \rangle$ be models with designated points and $r < \omega$. The $r$-round bisimulation game $\mathcal{G}(\mathcal{M}_1, \mathcal{M}_2, x, y)$ is played by two players, $\forall$ (Abelard, man) and $\exists$ (Heloise, woman). The rules of the game are the following:
  \begin{enumerate}
    \item ({\bf Round 0}) $\exists$ wins, if $x \equiv_0 y$. Otherwise, $\forall$ wins.
    \item ({\bf Round $n+1$}) Let $n + 1 < r$. Suppose rounds $0, \dots, n$ have been played.
    The $n$-th position is a pair $(x_n, y_n)$.
    $\forall$ chooses $x_{n + 1}$ (or $y_n \in W_2$) with $x_n R_1 x_{n + 1}$ (or $y_n R_1 y_{n + 1}$). $\exists$ chooses $y_{n + 1}$ ($x_{n + 1}$) with $y_n R_1 y_{n + 1}$ (or $y_n R_1 y_{n + 1}$) and $x_{n + 1} \equiv_0 y_{n + 1}$.
    $\forall$ ($\exists$) wins, if $\exists$ ($\forall$) cannot move.
  \end{enumerate}
  $\exists$ wins the game after $r$ rounds. $\forall$ wins the game, if there exists $k < r$ such that $\forall$ wins in {\bf Round $k$}.
\end{defin}

There are two $n$-equivalences for formulas and games (the first one is due to Fine \cite{fine1974logics}). These definitions are equivalent, this is due to Stirling.

  \begin{itemize}
  \item $(\mathcal{M}_1, x) \equiv_n (\mathcal{M}_2, y)$ iff for every $\varphi$ such that
  $md(\varphi) \leq n$ we have $\mathcal{M}_1, x \models \varphi$ iff $\mathcal{M}_2, y \models \varphi$.
  \item $(\mathcal{M}_1, x) \equiv_n (\mathcal{M}_2, y)$ iff $\exists$ has a winning strategy in $\mathcal{G}(\mathcal{M}_1, \mathcal{M}_2, x, y)$.
  \end{itemize}
\end{defin}

\begin{lemma} (Stabilisation lemma)

  Let $n$ be a natural number and $\mathcal{L}$ a modal logic.
  If $\equiv_n = \equiv_{n + 1}$ in a weak canonical model $\mathcal{M}_{\mathcal{L}} \upharpoonright k$, then $md(\mathcal{L} \upharpoonright k) \leq n$.
\end{lemma}

\begin{theorem} [See \cite{gabbay1998products}]
  Let $\mathcal{L}$ be a modal logic such that $md(\mathcal{L}) = n$ for some $n < \omega$, then $\mathcal{L}$ is locally finite.
\end{theorem}

@@

The ${\bf K5}$-closure (the Euclidean Horn closure of a binary relation) has the following equivalent definitions:
\begin{lemma} \label{equivHorn}
  Let $\mathcal{F} = \langle W, R \rangle$ be a Kripke frame and $R \subseteq R^{\bf K5}$.
  The following conditions are equivalent:

@@

Note that the following property holds not for every ${\bf K5}^{i,j}$, consider ${\bf KB}$ as an example (clearly, ${\bf KB} = {\bf K5}^{0,1}$), whose modal depth is $\infty$.

\begin{theorem}
$md({\bf K5}^{i,j}) = 1 + j + i$, if $i \neq 0$.
\end{theorem}

\begin{proof}

\end{proof}

@@

  \begin{enumerate}
  \item  $\mathcal{M} = \langle W, R, \vartheta \rangle$ be a ${\bf K}5$-model. Consider $\Gamma' = \Gamma \cup \{ \Diamond \Box \psi \: | \: \Box \psi \in \Gamma \}$. Let $\Delta$ be any finite and $\operatorname{Sub}$-closed extension of $\Gamma'$ that contains all $\Diamond \Box \psi$ for each $\Box \psi \in \Delta$. Then a model $\widehat{M} = \langle W / \sim_{\Delta}, (R^{min}_{\Delta})^{\bf K5}, \widehat{\vartheta} \rangle$ is a filtration of $\mathcal{M}$ through $\Delta$.
  \item Let $\mathcal{L} = {\bf K} \oplus \Diamond \Diamond \Diamond p \to \Diamond p$, then we have the similar statement for $\mathcal{L}$, where $\Delta$ is an extension of $\Gamma' = \{ \Diamond \Diamond \psi \: | \: \Diamond \psi \}$ that contains all $\Diamond \Diamond \varphi$ for $\Diamond \varphi \in \Delta$.
  \end{enumerate}
\end{lemma}
